\documentclass[11pt, one side, article]{memoir}


\settrims{0pt}{0pt} % page and stock same size
\settypeblocksize{*}{34pc}{*} % {height}{width}{ratio}
\setlrmargins{*}{*}{1} % {spine}{edge}{ratio}
\setulmarginsandblock{.98in}{.98in}{*} % height of typeblock computed
\setheadfoot{\onelineskip}{2\onelineskip} % {headheight}{footskip}
\setheaderspaces{*}{1.5\onelineskip}{*} % {headdrop}{headsep}{ratio}
\checkandfixthelayout

\usepackage{amsthm}
\usepackage{mathtools}
\usepackage{rotating}

\usepackage[inline]{enumitem}
\usepackage{ifthen}
\usepackage[utf8]{inputenc} %allows non-ascii in bib file
\usepackage{xcolor}
\usepackage{tikz-qtree}
\usepackage{subcaption}



\usepackage[bookmarks=true, colorlinks=true, linkcolor=blue!50!black, citecolor=orange!50!black, urlcolor=orange!50!black, pdfencoding=unicode]{hyperref}
\usepackage[capitalize]{cleveref}

\usepackage{tikz}

\usepackage{amssymb}
\usepackage{newpxtext}
\usepackage[varg,bigdelims]{newpxmath}
\usepackage{mathrsfs}
\usepackage{dutchcal}
\usepackage{mathalfa}
\usepackage{fontawesome}
\usepackage{ebproof}
\usepackage{stmaryrd}
\usepackage{ebproof}
\usepackage{graphicx}

% xcolor %
	\newcommand{\myred}[1]{{\color{red!60!black}#1}}
	\newcommand{\myyellow}[1]{{\color{yellow!60!black}#1}}
	\newcommand{\mygreen}[1]{{\color{green!40!black}#1}}

% cleveref %
  \newcommand{\creflastconjunction}{, and\nobreakspace} % serial comma
  \crefformat{enumi}{\card#2#1#3}
  \crefalias{chapter}{section}


% biblatex %

% hyperref %
  \hypersetup{final}

% enumitem %
  \setlist{nosep}
  \setlistdepth{6}


\include{style}

\newcommand{\dnote}[1]{{\quad \color{blue}$\lozenge$\;David says:}~#1\;{\color{blue}$\lozenge$}\quad}
\newcommand{\snote}[1]{{\quad \color{red}$\lozenge$\;Sophie says:}~#1\;{\color{red}$\lozenge$}\quad}
\newcommand{\defined}[1]{\textbf{#1}}



\newcommand{\thanksAFOSR}[1]{This material is based upon work supported by the Air Force Office of Scientific Research under award numbers #1}

% ---- Changeable document parameters ---- %

\linespread{1.1}
\allowdisplaybreaks
\setsecnumdepth{section}
\settocdepth{section}
\setlength{\parindent}{15pt}
\setcounter{tocdepth}{1}


\newcommand{\alice}{\textcolor{cyan}{\mathsf{alice}}}
\newcommand{\bob}{\textcolor{magenta}{\mathsf{bob}}}
\newcommand{\carmen}{\textcolor{teal}{\mathsf{carmen}}}
\newcommand{\aone}{\textcolor{cyan}{1}}
\newcommand{\azero}{\textcolor{cyan}{0}}
\newcommand{\bone}{\textcolor{magenta}{1}}
\newcommand{\bzero}{\textcolor{magenta}{0}}
\newcommand{\cone}{\textcolor{teal}{1}}
\newcommand{\czero}{\textcolor{teal}{0}}

\usepackage{forest}

\usepackage[backend=biber, backref=true, maxbibnames = 10, style = alphabetic]{biblatex}

\addbibresource{Library20240724.bib} 


%--------------- Document ---------------%
\begin{document}

\title{Dynamic task delegation for hierarchical agents}
% \title{Dynamic delegation for hierarchical agency}

\author{
	Sophie Libkind\and
	David I. Spivak
}

%\date{Last updated: \today}

\maketitle

\begin{abstract}
This is the fourth installment in a series of papers offering models of hierarchical structure for dynamical systems, using the language of polynomial functors. In the first iteration  \cite{niu2022poly}, the operad underlying the symmetric monoidal category $(\poly, \otimes, \yon)$ can be viewed as defining the behavior of hierarchical delegation. In particular, a morphism $\poly(p_1 \otimes \cdots \otimes p_m, q)$ turns the outputs of subordindates with interfaces $p_i$ into the output of an agent with interface $q$ and turns a task given to the agent into a task for each of the subordinates. The work of \cite{shapiro2022dynamic} extends this story by making it \emph{dynamic}. In other words, the way that subordinate outputs become agent outputs and agent tasks become subordinate tasks can change over time rather than remaining fixed. Most recently \cite{libkind2024pattern} further extends the story of hierarchical delegation in two different ways. First, it allows one to specify the hierarchical delegation programmatically (which we refer to as \emph{pattern}) rather than via behaviors (which we refer to as \emph{matter}).  Second, it allows for the time scales of the subordinates and agents to differ. An agent may rely on several outputs of its subordinates in order to produce a single output of its own.

In this article, we extend the story yet again. First, we extend the framework so that subordinates may be invoked asynchronously depending on the outcomes of other subordinates. From the perspective of programs/pattern, this extension takes the form of a $\smcat$-enriched operad $\org_\free$. From the perspective of behavior/matter, this extension takes the form of a $\smcat$-enriched operad $\org^\cofree$.  Second, we crispen the relationship between the programmatic and behavioral perspectives via a functor $[-, t] \colon \org_\free\op \to \org^\cofree$ for any polynomial monad $t$.

\end{abstract}

\input{0_intro}
\input{1_background}
\input{2_Org-m}
\input{3_Org-c}
\input{4_matter-pattern}


\chapter{Future work}

We can think of a few directions for future work. The first is simply to implement these ideas in working software or to find it modeled in living systems. For example, this theory could have applications in organizational design, robotics, biology, neuroscience, AI, ecology, etc. 

The second is to relax a certain ``closed world'' orientation that the present paper has. That is, the systems presented here focus on tasks that are hierarchically delegated and solved by subsystems of subsystems, possibly with an end user making final choices. It seems important to relax this to model a world in which the source of responses or inputs is somehow more extemporaneous.

The third is to consider behavior contracts for $\org_\free$, as in \cite{schultz2016dynamical,spivak2021learnersv1}. Each hom-category $\org_\free(p;q_1,\ldots,q_n)$ is a topos, and its internal language can be used to limit which dynamic hierarchical agents we want to consider. At a high level, this is mathematically straightforward using Kripke-Joyal semantics \cite[Chapter VI]{macLane1992sheaves}, however the sort of topos we're considering, toposes of polynomial coalgebras, are quite specific and thus have certain logical constructs that don't exists generically. Moreover, because we are interested in composing systems, we need to combine these logical constructs across several toposes, and an overarching logical framework would be useful.

\printbibliography

\end{document}